---
title: "Exercise 2"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(arrow)
```

### Sameer Kamal

## Load data

Load the following data:
  + applications from `app_data_sample.parquet`
  + edges from `edges_sample.csv`

```{r load-data, echo=FALSE}
# change to your own path!
data_path <- "/Users/sameerkamal/Desktop/MBA/McGill - Work/People Analytics/"
applications <- read_parquet(paste0(data_path,"app_data_sample.parquet"))

```

## Get gender for examiners

We'll get gender based on the first name of the examiner, which is recorded in the field `examiner_name_first`. We'll use library `gender` for that, relying on a modified version of their own [example](https://cran.r-project.org/web/packages/gender/vignettes/predicting-gender.html).

Note that there are over 2 million records in the applications table -- that's because there are many records for each examiner, as many as the number of applications that examiner worked on during this time frame. Our first step therefore is to get all *unique* names in a separate list `examiner_names`. We will then guess gender for each one and will join this table back to the original dataset. So, let's get names without repetition:

```{r gender-1,echo=FALSE}
library(gender)
#install_genderdata_package() # only run this line the first time you use the package, to get data for it

# get a list of first names without repetitions1
examiner_names <- applications %>% 
  distinct(examiner_name_first)

```

Now let's use function `gender()` as shown in the example for the package to attach a gender and probability to each name and put the results into the table `examiner_names_gender`

```{r gender-2, echo=FALSE}
# get a table of names and gender
examiner_names_gender <- examiner_names %>% 
  do(results = gender(.$examiner_name_first, method = "ssa")) %>% 
  unnest(cols = c(results), keep_empty = TRUE) %>% 
  select(
    examiner_name_first = name,
    gender,
    proportion_female
  )

```

Finally, let's join that table back to our original applications data and discard the temporary tables we have just created to reduce clutter in our environment.

```{r gender-3, echo=FALSE}
# remove extra colums from the gender table
examiner_names_gender <- examiner_names_gender %>% 
  select(examiner_name_first, gender)

# joining gender back to the dataset
applications <- applications %>% 
  left_join(examiner_names_gender, by = "examiner_name_first")

# cleaning up
rm(examiner_names)
rm(examiner_names_gender)
gc()

```


## Guess the examiner's race

We'll now use package `wru` to estimate likely race of an examiner. Just like with gender, we'll get a list of unique names first, only now we are using surnames.

```{r race-1, echo=FALSE}
library(wru)

examiner_surnames <- applications %>% 
  select(surname = examiner_name_last) %>% 
  distinct()

```
We'll follow the instructions for the package outlined here [https://github.com/kosukeimai/wru](https://github.com/kosukeimai/wru).

```{r race-2,echo=FALSE}
examiner_race <- predict_race(voter.file = examiner_surnames, surname.only = T) %>% 
  as_tibble()

```

As you can see, we get probabilities across five broad US Census categories: white, black, Hispanic, Asian and other. (Some of you may correctly point out that Hispanic is not a race category in the US Census, but these are the limitations of this package.)

Our final step here is to pick the race category that has the highest probability for each last name and then join the table back to the main applications table. See this example for comparing values across columns: [https://www.tidyverse.org/blog/2020/04/dplyr-1-0-0-rowwise/](https://www.tidyverse.org/blog/2020/04/dplyr-1-0-0-rowwise/). And this one for `case_when()` function: [https://dplyr.tidyverse.org/reference/case_when.html](https://dplyr.tidyverse.org/reference/case_when.html).

```{r race-3,echo=FALSE}
examiner_race <- examiner_race %>% 
  mutate(max_race_p = pmax(pred.asi, pred.bla, pred.his, pred.oth, pred.whi)) %>% 
  mutate(race = case_when(
    max_race_p == pred.asi ~ "Asian",
    max_race_p == pred.bla ~ "black",
    max_race_p == pred.his ~ "Hispanic",
    max_race_p == pred.oth ~ "other",
    max_race_p == pred.whi ~ "white",
    TRUE ~ NA_character_
  ))

```

Let's join the data back to the applications table.

```{r race-4,echo=FALSE}
# removing extra columns
examiner_race <- examiner_race %>% 
  select(surname,race)

applications <- applications %>% 
  left_join(examiner_race, by = c("examiner_name_last" = "surname"))

rm(examiner_race)
rm(examiner_surnames)
gc()
```


## Examiner's tenure 

To figure out the timespan for which we observe each examiner in the applications data, let's find the first and the last observed date for each examiner. We'll first get examiner IDs and application dates in a separate table, for ease of manipulation. We'll keep examiner ID (the field `examiner_id`), and earliest and latest dates for each application (`filing_date` and `appl_status_date` respectively). We'll use functions in package `lubridate` to work with date and time values.

```{r tenure-1,echo=FALSE}
library(lubridate) # to work with dates

examiner_dates <- applications %>% 
  select(examiner_id, filing_date, appl_status_date) 

```

The dates look inconsistent in terms of formatting. Let's make them consistent. We'll create new variables `start_date` and `end_date`.

```{r tenure-2,echo=FALSE}
examiner_dates <- examiner_dates %>% 
  mutate(start_date = ymd(filing_date), end_date = as_date(dmy_hms(appl_status_date)))
```

Let's now identify the earliest and the latest date for each examiner and calculate the difference in days, which is their tenure in the organization.

```{r tenure-3,echo=FALSE}
examiner_dates <- examiner_dates %>% 
  group_by(examiner_id) %>% 
  summarise(
    earliest_date = min(start_date, na.rm = TRUE), 
    latest_date = max(end_date, na.rm = TRUE),
    tenure_days = interval(earliest_date, latest_date) %/% days(1)
    ) %>% 
  filter(year(latest_date)<2018)

```

Joining back to the applications data.

```{r tenure-4,echo=FALSE}
applications <- applications %>% 
  left_join(examiner_dates, by = "examiner_id")

rm(examiner_dates)
gc()
```

## Visualizing gender, race and tenure distributions

```{r gg,echo=FALSE}
library(ggplot2)
```

```{r plotting,echo=FALSE}
ggplot(applications)+
  geom_bar(mapping = aes(x = gender), fill = "green")+
  ggtitle("Gender Distribution")+
  xlab("Gender")+
  ylab("Frequency")+
  theme_minimal()
ggplot(applications)+
  geom_bar(mapping = aes(x = race), fill = "purple")+
  scale_y_continuous(labels = scales::comma_format())+
  ggtitle("Race")+
  xlab("Race")+
  ylab("Frequency")+
  theme_minimal()
ggplot(applications, aes(x = tenure_days), na.rm = TRUE) +
  geom_histogram(binwidth = 100, fill = "blue") +
  ggtitle("Tenure") +
  xlab("Tenure") +
  ylab("Frequency")+
  theme_minimal()
```

## Visualizing distribution across Technology Centres

```{r plotting 2,echo=FALSE}

# Create a vector of labels
ttc <- applications$tc
labels <- unique(ttc)
x_labels <- paste0("TC", labels)

# Create a bar chart with custom x-axis labels
ggplot(applications, aes(x = tc)) + 
  geom_bar(fill = "red") +
  scale_x_continuous(breaks = labels, labels = x_labels)+
  scale_y_continuous(labels = scales::comma_format())+
  ggtitle("Distribution across technology centres") +
  xlab("Technology centre") +
  ylab("Frequency")+
  theme_minimal()
```

## Visualizing distribution across work groups

```{r plotting 3,echo=FALSE}
ggplot(applications, aes(x = examiner_art_unit)) +
  geom_histogram(binwidth = 10, fill = "pink") +
  scale_y_continuous(labels = scales::comma_format())+
  ggtitle("Distribution across work group") +
  xlab("Work Group") +
  ylab("Frequency")+
  theme_minimal()
```

## Correlating gender and race with tenure (excluding Technology Centre)
```{r corr,echo=FALSE}
model1 <- lm(formula = tenure_days ~ 1 + factor(gender) + factor(race), data=applications)

summary(model1)
```

## Correlating gender and race with tenure and including Technology Centre

```{r corr 2,echo=FALSE}
model2 <- lm(formula = tenure_days ~ 1 + factor(gender) + factor(race)+factor(tc), data=applications)

summary(model2)
```

## Findings
Based on the results above, we see that race and gender are statistically significant predictors of tenure at the patent office. For example, we see that men, on average, have a lower average tenure, and the same could be said of Hispanics when compared to other ethnicities. Looking at the distributions, however, we see that sample sizes vary considerably between genders and ethnicities. In addition, the low R-squared values demonstrate the model is not a good fit - there are either missing predictors or other predictors that better explain the model variation. 

When we add technology centres to the model, we see a significant change in the coefficient for gender and an overall improvement in the adjusted R-squared. The change in adjusted R-squared shows that adding the new variables has genuinely improved the model fit, while the considerable change in the gender coefficient show that predictions previously attributed to gender may have more to do with differences in technology centres. It seems likely that there are significant differences in gender ratios in different technology centres.